---
title: Ex situ Gap Analysis
author: Report generate from the [GAMMA application](https://atlantabg.shinyapps.io/GAMMA)
output:
  html_document:
      css: style.css
date: "date: `r Sys.Date()`"
params:
  points: NULL
  pointsBuffer: NULL
  ersex: NULL 
  grsex: NULL
  srsScore: NULL
  ersScore: NULL
  grsScore: NULL
  bufferDist: NULL
---

```{=html}
<style type="text/css">
.main-container {
  max-width: 2800px;
  margin-left: auto;
  margin-right: auto;
}
</style>
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load("leaflet","data.table", "tidyr", "dplyr","plotly","DT","RColorBrewer", "readr")

# source some global data 
source("utilities/loadGlobalVariables.R")
# adding paths to the image features 

### function for control number of character after the decimal place 
sigfig <- function(vec, n=2){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig

formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

} 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# processing steps for map 
points <- params$points
pointBuffer <- params$pointsBuffer
ersex <- params$ersex 
grsex <- params$grsex
srsScore <- params$srsScore
ersScore <- params$ersScore
grsScore <- params$grsScore
bufferDist <- params$bufferDist
# process points layers 
# spilt out the G and H points
# assign icon for visualization 
points <- points |>
  dplyr::mutate(
    # icon = dplyr::case_when(
    #   `Current Germplasm Type` == "G" & source == "GBIF" ~ icon_paths$g_gbif, # Use & for vectorized AND
    #   `Current Germplasm Type` == "H" & source == "GBIF" ~ icon_paths$h_gbif, # Use & for vectorized AND
    #   `Current Germplasm Type` == "G" & source == "upload" ~ icon_paths$g_upload, # Corrected source name
    #   `Current Germplasm Type` == "H" & source == "upload" ~ icon_paths$h_upload, # Corrected source name
    #   TRUE ~ "default_icon" # Optional: Default icon if no conditions match
    # ),
    color = case_when(
          `Current Germplasm Type` == "H" ~ combinedColor[1],
          `Current Germplasm Type` == "G" ~ combinedColor[2]
        )
  )
## G
gGap <- points |>
    dplyr::filter(`Current Germplasm Type` == "G")
gLabels <- lapply(gGap$popup, htmltools::HTML)
## H
hGap <- points |>
    dplyr::filter(`Current Germplasm Type` == "H")
hLabels <- lapply(hGap$popup, htmltools::HTML)


# h buffers
h_buffer <- terra::aggregate(
    pointBuffer[pointBuffer$`Current Germplasm Type` == "H", ]
  )
#G buffers  
g_buffer<- terra::aggregate(
    pointBuffer[pointBuffer$`Current Germplasm Type` == "G", ]
  )

# generate the written summary 
srs <- as.numeric(srsScore) |> round(digits = 2)
grs <- as.numeric(grsScore) |> round(digits = 2)
ers <- as.numeric(ersScore) |> round(digits = 2)
fcs <- mean(c(srs,grs,ers), na.rm = TRUE) |> round(digits = 2)
totalPoints <- nrow(points)
referncePoints <- nrow(points[points$`Current Germplasm Type` == "H",])
germplasmPoints <- nrow(points[points$`Current Germplasm Type` == "G",])
species <- points$`Taxon Name`[1]

# generate the figure 
# define the base table
plotDF <- data.frame(class = c(
  "Sampling Representativeness Score",
  "Ecological Representativeness Score",
  "Geographic Representativeness Score",
  "Final Representativeness Score"),
  score = c(0,0,0,0))

# assign values based on the presence of specific output values
plotDF$score[1] <- srs
plotDF$score[2] <- ers
plotDF$score[3] <- grs
# assign the fcsex score
plotDF$score[4] <- fcs

# assign color based on the score
plotDF <- plotDF|>
  dplyr::mutate(rank = case_when(
    score <= 25 ~ "Urgent Priority",
    score > 25 & score <= 50 ~ "High Priority",
    score > 50 & score <= 75 ~"Medium Priority",
    score > 75 ~ "Low Priority"
  ),
  colors = case_when(
    rank == "Urgent Priority" ~ "#ffb4b3",
    rank ==  "High Priority" ~ "#ffd380",
    rank == "Medium Priority"~ "#ffff80",
    rank == "Low Priority" ~ "#a8d2a8"
  )
)

```

# Summary of results for `r  species`

The gap analysis was conducted using a total of `r totalPoints` records. Of these `r germplasmPoints` were germplasm records and `r referncePoints` were reference records.
<br> 

The relationship between these observation types is recorded by a sampling representativeness score. 
<br> 
A buffer size of `r bufferDist` was used to generate the results for the ecological and geographic representativeness score. 

The average for these three scores is used to calculate a final exsitu conservation score. 

**Sampling Representativeness Score**: `r srs`.
<br>
**Geographic representativeness score**: `r grs`
<br>
**Ecological representativeness score**: `r ers`.
<br>
**Final conservation score**: `r fcs`. 

Definitions of ex situ gap analysis scores are below.


# Map

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# inital map for gap analysis page
map2 <-  leaflet(options = leafletOptions(minZoom = 3, maxZoom = 16)) |>
    addProviderTiles("OpenStreetMap", group = "OpenStreetMap") |>
    addProviderTiles("Esri.WorldTopoMap", group = "Topography") |>
    addProviderTiles("Esri.WorldImagery", group = "Imagery")|>
    addLayersControl(
      position = "topleft",
      overlayGroups = c("Reference Records",
                        "Germplasm Records",
                        "Buffers",
                        "GRS gaps",
                        "ERS gaps"),
      baseGroups = c("OpenStreetMap",
                     "Topography",
                     "Imagery"),
      options = layersControlOptions(collapsed = FALSE)
    )|>
    hideGroup(c("GRS gaps","ERS gaps")) |> # keeping enabled for now
    addMapPane("ersGap", zIndex = 400) |>  # Lower zIndex for polygons
    addMapPane("grsGap", zIndex = 405) |>
    addMapPane("allBuffers", zIndex = 410) |>
    addMapPane("pointsH", zIndex = 415) |>   # Higher zIndex for points1
    addMapPane("pointsG", zIndex = 420)

# test for and add H points 
if(nrow(hGap) > 0){
  map2 <- map2 |>
        # add point objects
    addCircleMarkers(
        data = hGap,
        group = "Reference Records",
        radius = 4,
        color = "white",
        fillColor = ~color,
        stroke = TRUE,
        weight = 1,
        fillOpacity = 1,
        label = hLabels
      )|>
      # add buffer objects
      addPolygons(data = h_buffer,
                  group = "Buffers",
                  color = combinedColor[1],
                  fillOpacity = 0.5,
                  options = pathOptions(pane = "allBuffers"))
}
# test for and add H points 
if(nrow(gGap) > 0){
  map2 <- map2 |>
    # add point objects
    addCircleMarkers(
        data = gGap,
        group = "Germplasm Records",
        radius = 4,
        color = "white",
        fillColor = ~color,
        stroke = TRUE,
        weight = 1,
        fillOpacity = 1,
        label = gLabels
      )|>
    # add buffers 
      addPolygons(data = g_buffer,
              group = "Buffers",
              color = combinedColor[2],
              fillOpacity = 0.5,
              options = pathOptions(pane = "allBuffers"))
}
map2 <- map2 |>
      #add grs map
      clearGroup("GRS gaps")|>
      addPolygons(data = grsex,
                  group = "GRS gaps",
                  color = grsexColor,
                  fillOpacity = 0.5,
                  options = pathOptions(pane = "grsGap")) |>
      # add ersex map
      addPolygons(data = ersex_missingEcos(),
                  group = "ERS gaps",
                  color = ersexColor,
                  fillOpacity = 0.5,
                  popup = ~ECO_NAME,
                  options = pathOptions(pane = "ersGap"))|>
    addLegend(
        position = "bottomleft",
        # group = "GRS gaps",
        colors = combinedColor,
        labels = c("Reference", "Germplasm")
      )|>
      addLegend(
        position = "bottomleft",
        group = "GRS gaps",
        colors = grsexColor,
        labels = c("GRS gaps")
      )|>
      addLegend(
        position = "bottomleft",
        group = "ERS gaps",
        colors = ersexColor,
        labels = c("ERS gaps")
      )


map2
```

# Gap Analysis Scores 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
#define the display order of the plot
xform <- list(categoryorder = "array",
              categoryarray = c("Sampling Representativeness Score",
                                "Ecological Representativeness Score",
                                "Geographic Representativeness Score",
                                "Final Representativeness Score"))

# generate a plotly figure
fig <- plot_ly(
  data = plotDF,
  x = ~class,
  y = ~score,
  marker = list(color = c(plotDF$colors)),
  type = "bar"
)|>
  layout(title = "Gap Analysis Ex Situ Conservation Summary",
        xaxis = xform,
        yaxis = list(title = "",
                     range = c(0,100))
        )
# print figure
fig
    
```


# Table of gap analysis Data

The following table contains all the records that were used to generate the gap analysis results.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
colSelect <- points |>
  as.data.frame()|>
  dplyr::select(
    `Accession Number`,
    `Taxon Name`,
    `Current Germplasm Type`,
    `Collection Date` ,
    `source`,
    `Locality`,
    `Collector`,
    `Latitude` ,
    "Longitude"
  )

DT::datatable(colSelect,
              rownames = FALSE)
```

### Definitions of occurrence data categories
       
**Germplasm Records (G)** : Occurrences in which a living sample (via plant or seed) is present in an (*ex situ*), conservation system (i.e., botanical garden, seed bank, genebank, etc.).

**Reference Records (H)** : Occurrences that have a supporting herbarium or other reference record.


### Sampling Representativeness Score (SRS)

**Ex situ**: The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher (H) records for each taxon.

### Geographic Representativeness Score (GRS)

**Ex situ**: The Geographic Representativeness Score *ex situ* (GRS ex situ) uses a user defined km-radius buffer created around each G collection coordinate point to estimate geographic areas already well collected within the distribution of each taxon, also created using buffers around H reference points. This is calculated as the proportion of the distribution covered by the G buffers.


### Ecological Representativeness Score (ERS)

**Ex situ**: The Ecological Representativeness Score ex situ (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the potential distribution.

### Final Conservation Score (FCS)

**Ex situ**: The Final Conservation Score *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics.

**Prioritization using FSC**

In considering the analysis of multiple species, FSC may be used to aid prioritize species action with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥75.

