---
title: Ex situ conservation assessment 
author: GAMMA application 
output:
  html_document:
      css: style.css
date: "Report generated on `r Sys.Date()`"
params:
  reportData: NA
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
pacman::p_load("leaflet","data.table", "tidyr", "dplyr","plotly","DT","RColorBrewer")

### function for control number of character after the decimal place 
sigfig <- function(vec, n=2){ 
### function to round values to N significant digits
# input:   vec       vector of numeric
#          n         integer is the required sigfig  
# output:  outvec    vector of numeric rounded to N sigfig

formatC(signif(vec,digits=n), digits=n,format="fg", flag="#") 

} 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# transform data

testing <- FALSE
if(testing == TRUE){
  data <- reportData # testing only
  species <- reportData$countsData$species # testing
}else{
  data <- params$reportData
  species <- data$occuranceData$taxon[1]
}
printSpecies <- stringr::str_replace_all(string = species, pattern = "_", replacement = " ")
```

# `r printSpecies`
<br>
## Occurrence records    

The table below shows a summary of the occurrence records for the taxon. A record “with Lat/Long” is one that has a complete set of latitude and longitude values associated with it, which could be used for distribution modeling and the conservation gap analysis. In preparation for the conservation gap analyses, we classified each record based on whether it was an existing ex situ germplasm sample from a genebank, botanic garden, or other repository (labeled G, as most records were from genebanks), or a reference observation/voucher specimen (labeled H, as most of these records were from herbaria).

```{r  echo=FALSE, message=FALSE, warning=FALSE}
# alter Counts data for better visualizations
c1 <- data$countsData %>%
  dplyr::select(
    Taxon = species,
    'Occurrences' = totalRecords,
    'Occurrences with Lat/Long' = totalUseful,
    'Germplasm Records(G)' = totalGRecords,
    'Reference Records(H)' = totalHRecords,
    'Unique Data Sources' = numberOfUniqueSources
  ) %>%
  dplyr::mutate(Taxon = stringr::str_replace_all(Taxon, pattern = "_", replacement = " "))


DT::datatable(c1,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```


```{r, echo=FALSE, eval=TRUE}
# build condition to test for NA in data$modelEvalation 
```


<br>


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Generate all the map data
# point data
occData <- data$occuranceData %>%
  dplyr::mutate(
    popup = paste0(
      "<br/><b>Taxon:</b> ",
      taxon,
      "<br/><b>Source:</b> ",
      databaseSource,
      "<br/><b>Collector Code:</b> ",
      institutionCode,
      "<br/><b>Collection Type:</b> ",
      type
    )
  ) |>
  mutate(color = case_when(type == "H" ~ "#1184D4",
                           type == "G" ~ "#6300F0"))

webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project

if(data$NoModel == FALSE){
    # predicted presence map
  threshold <- data$binaryMap
  thresColor <- colorNumeric(
    palette = c("#cfcccc70", "#45B320"),
    domain = values(threshold),
    alpha = TRUE,
    na.color = "transparent"
  )
  # generate a threshold mask
  mask1 <- ifel(test = threshold == 1, yes = 1, no = NA)
  # protected area
  protectedArea <- data$protectedArea
  protectColor <-
    colorNumeric(
      palette = "#B7B648",
      domain = values(protectedArea),
      alpha = TRUE,
      na.color = "transparent"
    )
  # protected areas mask
  proMask <- ifel(test = protectedArea == 1, yes = 1, no = NA)
   # insitu map elements
  ## SRSIN
  ## total number of observations per cell of the threshold model masked against the protect areas layer 
  # filter points outside of protected areas 
  webMecOccData <- vect(st_transform(occData, crs = "epsg:3857")) # need to change project
  # for the spatial operations with the layer. occData can't have the proj
  # get all the points that are outside of a protected area
  
  
  inThres <- terra::extract(x = mask1, y = webMecOccData)|>
    dplyr::filter(Threshold==1)
  inPro <- terra::extract(x = protectedArea, y = webMecOccData) |>
    dplyr::filter(is.na(layer))

  srs1 <- terra::rasterize(webMecOccData,
                     threshold,
                     fun = "sum",
                     background = 0) |>
    mask(mask1)
  
  totalObsinThres <- nrow(inThres)
  ## total number of observations in protected areas
  totalObsinPro <- nrow(inThres) - nrow(inPro)

  ## GRSIN
  ## reclassy protected area
  proReclass <- ifel(test = is.na(protectedArea),
                     yes = 0,
                     no = 1)
  ## add values
  grs1 <- threshold + proReclass
  exPal <- c("#cfcccc80", "#45b320", "#7570b3")
  # for some reason passing the vector exPal to the palette arg does not maintain the alpha reference? 
  grsinColor <- colorNumeric(palette = c("#cfcccc80", "#45b320", "#7570b3"),
                 # purple is a little lighter so the point locations show up clearly
                 domain = values(grs1),
                 alpha = TRUE,
                 na.color = "transparent")
  ## ERSin
  # used on the ERS visualization
  allEcos <- data$naturalArea |>
    dplyr::select("ECO_ID_U", "ECO_NAME")
  # reproject for spatial operations
  webMecAllEcos <- st_transform(allEcos, crs = "epsg:3857")
  ## showcase ecoregions within the threshold model that do no have a protected area
  ers1 <- terra::zonal(proReclass,
                 vect(webMecAllEcos),
                 fun = "sum",
                 na.rm = TRUE)
  ers1$ECO_ID_U <- allEcos$ECO_ID_U
  ers1$ECO_NAME <- allEcos$ECO_NAME
  
  ## zonal Stats to determine overlap between the ecoregions and the threshold model
  v1 <- terra::zonal(x = threshold,
                 z = vect(webMecAllEcos),
                 fun = "sum",
                 na.rm = TRUE)# , exact = TRUE)

  ## reassign the eco id record
  v1$ECO_ID_U <- webMecAllEcos$ECO_ID_U
  # get a list of ecoregions within the threshold model
  ecosInThres <- v1 |>
    dplyr::filter(Threshold > 0) |>
    dplyr::select(ECO_ID_U) |>
    pull()
  
  ## ID ecoregions in threshold model with no protected areas
  ers2 <- ers1 |>
    dplyr::filter(ECO_ID_U %in% ecosInThres) |>
    dplyr::filter(layer == 0)

if(nrow(ers2) > 0) {
  ## use this list to generate a raster object masked to the
  ersinMap <- webMecAllEcos |>
    dplyr::filter(ECO_ID_U %in% ers2$ECO_ID_U) |>
    terra::rasterize(y = threshold, "ECO_ID_U") |>
    mask(mask = mask1)
  # generate palatte object for map 
  ecoPal2 <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
  ecoPal2 <- ecoPal2[1:length(ers2$ECO_ID_U)]
  # need to order the names so they appear as expected 
  ers2$color <- ecoPal2
  ers2 <- ers2 |> arrange(desc(ECO_NAME))
  ers2$color <- ecoPal2
  
  ers2$ECO_ID_U <- factor(ers2$ECO_ID_U , ordered = TRUE, 
                                  levels = c(ers2$ECO_ID_U))
  
  
  # define a palette option so that we can set a transparent NA value
  ecoInColor <- colorFactor(palette =ers2$color,
                            domain = ers2$ECO_ID_U,
                            na.color =  "#FF000000")
  
    
} else{
  ersinMap <- "No values"
}



# exsitu map elements

#buffered G area
gBuf <- data$g_bufferCrop
if (class(gBuf) != "character" & data$fcsex$GRS != 0) {
  bufColor <-
    colorNumeric(
      palette = "#6300F0",
      domain = values(gBuf),
      na.color = "transparent"
    )
  # GRSex map element
  m1 <- cbind(NaN, 0)
  gbuf2 <- terra::classify(x = gBuf, m1)
  grsexMap <- threshold + gbuf2
  # need to defined this outside of the palette function so it can be called again in the legend
  grsexColor <- colorNumeric(palette = c("#cfcccc80", "#45b320", "#7570b3"),
                 # purple is a little lighter so the point locations show up clearly
                 domain = values(grsexMap),
                 alpha = TRUE,
                 na.color = "transparent")
  }
  
  
  #ecoregions
  ## grab the missing ecoregions
  missingEcos <- data$ersex$missingEcos |> unlist()
  ## filter the full set
  ecoReg <- data$naturalArea |>
    dplyr::filter(ECO_ID_U %in% missingEcos) |>
    dplyr::mutate(ECO_NAME = as.factor(ECO_NAME))
  
  # need to generate a raster object ecoregions, this should be done as part of the modeling process
  # then I can pass through a data frame of the ecoregion data, rather the spatial objects
  
  # Rasterize the ecoregions
  webMecEcoReg <- st_transform(ecoReg,crs = 3857)
  ecoRast <- terra::rasterize(x = webMecEcoReg, y = threshold, "ECO_ID_U")
  
  # apply the mask
  ecoRast <- mask(ecoRast, mask1)
  
  
  # set the palette
  # this defines the color used, and can be called in the legend creation
  ## set 3 has a cap of 12 unique colors
  ### duplicating the values then subsetting base on the required number. 
  ecoPal <-  rep(RColorBrewer::brewer.pal(n = 12, name = "Spectral"), 6)
  ecoPal <- ecoPal[1:length(ecoReg$ECO_ID_U)]
  
  ecoReg$color <- ecoPal
  ecoReg <- ecoReg |> arrange(desc(ECO_NAME))
  ecoReg$ECO_ID_U <- factor(ecoReg$ECO_ID_U, ordered = TRUE, 
                                  levels = c(ecoReg$ECO_ID_U))
  
  # define a palette option so that we can set a transparent NA value
  ecoColor <- colorFactor(palette = ecoReg$color,
                          domain = ecoReg$ECO_ID_U,
                          na.color =  "#FF000000")
  
  
  #distribution palette
  distColor <- c("#ffffd4", "#fed98e", "#fe9929", "#d95f0e", "#993404")
  # protect area palette
  proColor <- c("#7570b3")
  # srsin palete
  srsPal1 <-  c("#cfcccc95",
      "#7fcdbb",
      "#41b6c4",
      "#1d91c0",
      "#225ea8",
      "#0c2c84")
  srsPal <- colorNumeric(palette = c("#cfcccc95","#7fcdbb","#41b6c4","#1d91c0","#225ea8","#0c2c84"),
                 domain = values(srs1),
                 alpha = TRUE,
                 na.color = "transparent")
  
  # Mean distribution map
  meanDist <- data$projectedResults$mean
  meanPal <- colorNumeric(palette = distColor,
                 domain = values(meanDist),
                 na.color = "transparent")
  # Median distribution map
  medianDist <- data$projectedResults$median
  medianPal <-
    colorNumeric(palette = distColor,
                 domain = values(medianDist),
                 na.color = "transparent")
  
  # standdard devation map
  stdevDist <- data$projectedResults$stdev
  stdevPal <- colorNumeric(palette = distColor,
                 domain = values(stdevDist),
                 na.color = "transparent")
}





#
```


<br>
## Map of Predicted distribution

The map below shows the median, mean, and standard deviation (SD) of potential distribution model runs, as well as the predicted presence threshold distribution. The predicted presence threshold distribution was generated by applying a statistically determined threshold value to the median model distribution, and is used in the conservation gap analysis as the potential distribution model. The occurrences used in the modeling process are represented by dots, distinguishing between ex situ germplasm collections (G) and reference sightings/voucher specimens (H, for herbaria) records (as available), and the attribute data associated with the points can be viewed by clicking on the dots. The semi-transparent gray background reflects the native country-ecoregion area.

Layers can be added to the map using the toggle buttons on the left side. The first layer turned on will be in the bottom position on the map.


```{r echo=FALSE, message=FALSE, warning=FALSE}
# modeled distribution Map
## Add occurrances 
m1 <- leaflet(width = "100%", options = leafletOptions(minZoom = 3)) |>
  addTiles()|>
  addCircleMarkers(data = occData[occData$type=="H",],
             color = ~color,
             opacity = 1,
             radius = 1,
             group = "Occurrences",
             stroke = 1,
             popup = ~popup) |> 
  addCircleMarkers(data = occData[occData$type=="G",],
             color = ~color,
             opacity = 1,
             radius = 1,
             group = "Occurrences",
             stroke = 1,
             popup = ~popup)|>
    leaflet::addLegend(
      position = "bottomright",
      color = c("#1184D4","#6300F0"),
      labels = c("H (for herbaria) observations", "G (for genebank) samples"),
      title = "Observation Type",
      group = "Occurrences")



if(data$NoModel == FALSE){
  ## add rasters 
m2 <- m1 |> 
  # threshold 
  addRasterImage(x = raster(threshold),
                 colors = thresColor, 
                 group = "Potential distribution model",
                 project = FALSE)|>
    leaflet::addLegend(position = "bottomright",
            color = c( "#cfcccc","#45B320"),
            labels = c("Native Area", "Potential distribution model"),
            title = "Modeled Habitat",
            group = "Potential distribution model")|>
  #mean 
  addRasterImage(x = meanDist, 
                 colors = meanPal,
                 group = "Mean Distribution Model",
                 project = FALSE)|>
    leaflet::addLegend(position = "bottomright",
            color = distColor,
            title = "Mean Distribution Model",
            labels = c("low","","","","high"),
            group = "Mean Distribution Model")|>
  #median 
  addRasterImage(x = medianDist, 
                 colors = medianPal,
                 group = "Median Distribution Model",
                 project = FALSE) |>
    leaflet::addLegend(position = "bottomright",
            color = distColor,
            title = "Median Distribution Model",
            labels = c("low","","","","high"),
            group = "Median Distribution Model")|>
    #stdev 
  addRasterImage(x = stdevDist, 
                 colors = stdevPal,
                 group = "SD of Distribution Model",
                 project = FALSE)|>
    leaflet::addLegend(position = "bottomright",
            color = distColor,
            title = "SD of Distribution Model",
            labels = c("low","","","","high"),
            group = "SD of Distribution Model")

## Add control layers 
m3 <- m2 |>
  addLayersControl(position = "topleft", options = list(collapsed = FALSE,
                                                        autoZIndex = FALSE),
    overlayGroups = c("Potential distribution model","Occurrences",
                      "Mean Distribution Model",
                      "Median Distribution Model",
                      "SD of Distribution Model"))|> 
  hideGroup(c("Mean Distribution Model",
                  "Median Distribution Model",
                  "SD of Distribution Model"))
m3 
}else{
  m1
}


```

<br>

### Evaluation Metrics 

For each taxon with at least 25 distinct coordinates in the target spatial area, the modeled distribution was calculated as the median of ten MaxEnt model replicates (K = 10), using linear, quadratic, product, and hinge features, with a regularization parameter ß=1.0. For taxa with 8-24 distinct coordinates, the median of six replicates (K = 6) was calculated, and for taxa with 3-7 distinct coordinates, the median of three replicates (K = 3) was calculated. Model runs for taxa where MaxEnt failed within these established groupings were rerun in fewer replicate categories. Taxa with fewer than 3 distinct coordinates were not modeled. The table below shows the statistical results for each replicate.


```{r, echo=FALSE, eval=TRUE}
# set the sigfig in the table

if(data$NoModel == FALSE){
  eval1 <- data$modelEvaluation %>%
  dplyr::select(
    "AUC" = "AUCtest" ,
    "Normalized AUC" = "nAUC",
    "Calibrated AUC" = "cAUC",
    "Threshold Value" = "threshold_test",
    "Sensitivity"  = "sensi_test",
    "Specificity" = "speci_test",
    "Mathews Correlation" = "matthews.cor_test",
    "Cohen's kappa" = "kappa_index_test"
  ) %>%
  dplyr::mutate_all(.funs = sigfig)


DT::datatable(eval1,
              class = "hover",
               options = list(dom = 't'))
}else{
  "No model was generated for this species"
}

```

<br>
### Predictors used in modeling process 

The table below shows the ecogeographic variables listed in order of importance based on the occurrence dataset, including whether or not they were used for distribution modeling, following the variable selection process. Bioclimatic predictors are from [WorldClim2](https://www.worldclim.com/version2). Topographic, wind speed and solar radiation predictors were generated by [CIAT DAPA](https://github.com/CIAT-DAPA) and can be made available upon request. 




```{r echo=FALSE, message=FALSE, warning=FALSE}

if(data$NoModel == FALSE){
  varTable <- data$variableImportance |> 
    dplyr::mutate(across(c('importance'), round, 3))|>
    dplyr::select("Predictor variable name" = `Current title`,
                "Relative importance to model" =  importance,
                "Included in the modeling process" = includeInFinal,
                "Abbrevaition" = varNames
                )|>
  dplyr::arrange(desc(`Relative importance to model`))|>
  dplyr::arrange(desc(`Included in the modeling process`))
# display table 
DT::datatable(varTable,
              rownames = FALSE,
              class = "compact")
}else{
  "No model was generated for this species"
}


```

<br>
<br>

#  *Ex situ* conservation gap analysis 

The table below shows the *ex situ* conservation gap analysis summary. The sampling representativeness score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher (H) records for each taxon, making use of all compiled records, regardless of whether they include coordinates. The geographical representativeness score *ex situ* (GRS ex situ) uses buffers of 50 km radius created around each G collection coordinate point to estimate geographic areas already well collected within the distribution models of each taxon, and then calculates the proportion of the distribution model covered by these buffers. The ecological representativeness score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the distribution model. A final conservation score for *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS *ex situ* is used to categorize species, with urgent priority (UP) for further conservation action assigned when FCS ex situ < 25, high priority (HP) where 25 ≤ FCS ex situ < 50, medium priority (MP) where 50 ≤ FCS ex situ < 75, and low priority (LP) for taxa whose FCS ex situ ≥75


```{r echo=FALSE, message=FALSE, warning=FALSE}
exCounts <- data$countsData |> 
  dplyr::mutate(across(c('species'), str_replace_all, pattern = "_", replacement = " "))|>
  dplyr::select("Taxon" = species, 
                "Total with coordinates" = totalUseful,
                "Total G records" = totalGRecords,
                "Total G with coordinates" = totalGUseful,
                "Total H records" = totalHRecords,
                "Total H with coordinates" = totalHUseful)
exScores <- data$fcsex |> 
  dplyr::mutate("SRS ex situ" = sigfig(SRS),
                "GRS ex situ" = sigfig(GRS),
                "ERS ex situ" = sigfig(ERS),
                "FCS ex situ" = sigfig(FCS))|>
  dplyr::select("SRS ex situ","GRS ex situ","ERS ex situ","FCS ex situ")|>
  dplyr::mutate(
    "FCS ex situ priority category" = case_when(
      `FCS ex situ` >= 75 ~ "LP",
      `FCS ex situ` >=50 & `FCS ex situ` < 75 ~ "MP",
      `FCS ex situ` >=25 & `FCS ex situ` < 50 ~ "HP",
      `FCS ex situ` < 25 ~ "UP"
    )
  )
# combine the data
exTables <- bind_cols(exCounts,exScores)
#visualize 
DT::datatable(exTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```
<br>
## Map of *ex situ* conservation gap analysis measures

The GRS ex situ map feature shows the potential distribution model, with previous *ex situ* germplasm collection points surrounded by a 50 km buffer overlaid. Only germplasm (G) points are displayed on the map.

The ERS ex situ map feature shows ecoregions within the potential distribution model from which no ex situ germplasm collections have been made. The ecoregion extent has been clipped to be within the potential distribution model of the taxon.There are many cases where the ecoregion area within the potential distribution model extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS ex situ legend to help locate these areas.


```{r echo=FALSE, message=FALSE, warning=FALSE}

if(data$NoModel == FALSE){
  
if(exTables$`Total G with coordinates` == 0){
  # generate the GRSex map
ex2 <- leaflet(width = "100%")|>
  addProviderTiles(providers$Esri.WorldGrayCanvas)|>
  # Outline of all ecoregions
  addPolygons(
    data = allEcos,
    group = "All ecoregions",
    weight = 2,
    stroke = TRUE,
    fill = TRUE,
    fillOpacity = 0.2,
    popup = allEcos$ECO_NAME,
    color = "#60616050"
  ) 
}else{ # G points with valid lat lon are present
  # generate the GRSex map
  ex1 <- leaflet(width = "100%")|>
  addProviderTiles(providers$Esri.WorldGrayCanvas)|>
  #GRSex raster
  addRasterImage(
    x = raster(grsexMap),
    colors = grsexColor,
    group = "GRS ex situ",
    project = FALSE) |>
  #G points
  addCircleMarkers(
    data = occData[occData$type == "G",],
    color = ~ color,
    opacity = 1,
    radius = 1,
    group = "GRS ex situ",
    stroke = 1,
    popup = ~ popup
  ) |>
  #add legend
  leaflet::addLegend(
    position = "bottomright",
    color = exPal,
    title = "Ex Situ geographical representativeness",
    labels = c("Native country-ecoregion area", "Potential distribution model", "Areas < 50 km from an ex situ collection"),
    group = "GRS ex situ"
  )
  
# generate the ERSex map
if(ersex$ERS == 100){
  ex2 <- ex1 |>
    # Outline of all ecoregions
  addPolygons(
    data = allEcos,
    group = "All ecoregions",
    weight = 2,
    stroke = TRUE,
    fill = TRUE,
    fillOpacity = 0.2,
    popup = allEcos$ECO_NAME,
    color = "#60616050"
  ) 
}else{
  ex2 <- ex1 |>
  # Ecoregion Raster Layers
  addRasterImage(raster(ecoRast),
               group = "ERS ex situ",
               colors = ecoColor,
               project = FALSE)|>
  leaflet::addLegend(colors = ecoReg$color,
            group = "ERS ex situ",
            labels = ecoReg$ECO_NAME) |>
  # Outline of all ecoregions
  addPolygons(
    data = allEcos,
    group = "All ecoregions",
    weight = 2,
    stroke = TRUE,
    fill = TRUE,
    fillOpacity = 0.2,
    popup = allEcos$ECO_NAME,
    color = "#60616050"
  ) 
}
}



# add controls for the layout
ex3 <- ex2 |>
  addLayersControl(
    position = "topleft",
    options = list(collapsed = FALSE,
                   autoZIndex = FALSE),
    overlayGroups = c(
      "GRS ex situ",
      "ERS ex situ",
      "All ecoregions"
    )
  ) |>
  hideGroup(c(
    "ERS ex situ","All ecoregions"
  ))

## print the map
ex3
}else{
  "No model was generate for this species "
}

```

<br>
### Ex situ conservation ecoregions table


The table below shows all the ecoregions utilized within the study.

The column ERS ex situ gap ecoregion denotes (True) if an ecoregion overlaps with the potential distribution model but is farther than 50 km from any ex situ conservation occurrence, and is thus considered a gap in ex situ conservation

```{r echo=FALSE, message=FALSE, warning=FALSE}

# # process some area measure of missing ecoregions 
### this is going to take a lot of conditional testing so I'm not sure I want it in now.
### with the eco regions visualized properly I think it's ok without. 
# aveCellSize <- mean(values(cellSize(ecoRast,unit="km")))
# misArea <- ecoRast |>
#   terra::freq() |>
#   dplyr::mutate(area = sigfig(count * aveCellSize, n=3))|>
#   dplyr::select(value, area)

# All ecoregions considered
# identify if they
## have modeled habitat
## are within a 50km buffer of a germplasm collection
if(data$NoModel == FALSE){
  eco2 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERSex Gap Ecoregion' = case_when(ECO_ID_U %in% missingEcos ~ TRUE,
                                      TRUE ~ FALSE)
  ) |>
  # dplyr::left_join(misArea, by = c("ECO_ID_U"="value"))|>
  dplyr::select(
    `ERS ex situ gap ecoregion` = 'ERSex Gap Ecoregion',
    # "Square KM missing" = area,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion ame` = "ECO_NAME"  ) |>
  arrange(desc(`ERS ex situ gap ecoregion`))
#visualize
DT::datatable(eco2,
              rownames = FALSE,
              class = "compact")

}else{
  eco2 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERSex Gap Ecoregion' =  TRUE) |> 
  # dplyr::left_join(misArea, by = c("ECO_ID_U"="value"))|>
  dplyr::select(
    `ERS ex situ gap ecoregion` = 'ERSex Gap Ecoregion',
    # "Square KM missing" = area,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion name` = "ECO_NAME"  ) |>
  arrange(desc(`ERS ex situ gap ecoregion`))
#visualize
  DT::datatable(eco2,
              rownames = FALSE,
              class = "compact")
}

``` 

<br>
<br>

# *In situ* conservation gap analysis

The table below shows the *in situ* conservation gap analysis summary. The sampling representativeness score *in situ* (SRS in situ) calculates the proportion of all occurrences of a taxon within its predicted distribution that fall within a protected area. The geographical representativeness score *in situ* (GRS in situ) compares the area (in km2) of the potential distribution model located within protected areas versus the total area of the model. The ecological representativeness score *in situ* (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the potential distribution model. A final conservation score for *in situ* (FCS in situ) was derived by calculating the average of the three *in situ* conservation metrics. All of the conservation metrics are on a scale from 0-100, with 0 = poor state of conservation and 100 = comprehensive conservation. The FCS in situ is used to categorize taxa, with urgent priority (UP) for further conservation action assigned when FCS in situ < 25, high priority (HP) where 25 ≤ FCS in situ < 50, medium priority (MP) where 50 ≤ FCS in situ < 75, and low priority (LP) for taxa whose FCS in situ ≥75.

```{r echo=FALSE, message=FALSE, warning=FALSE}

inCounts <- data$countsData |> 
  dplyr::mutate(across(c('species'), str_replace_all, pattern = "_", replacement = " "))|>
  dplyr::mutate("Total occurrences in protected areas" = round(data$countsData$totalUseful * (data$fcsin$SRS/100))) |> 
  dplyr::select("Taxon" = species, 
                "Total occurrences" = totalUseful,
                  "Total occurrences in protected areas")
if(data$NoModel == FALSE){
  inScores <- data$fcsin |> 
  dplyr::mutate("SRS in situ" = sigfig(SRS),
                "GRS in situ" = sigfig(GRS),
                "ERS in situ" = sigfig(ERS),
                "FCS in situ" = sigfig(FCS),
                "Total occurrences in modeled area" = totalObsinThres,
                "Total occurrences in protected areas" = totalObsinPro)|>
  dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
  dplyr::mutate(
    "FCS in situ priority category" = case_when(
      `FCS in situ` >= 75 ~ "LP",
      `FCS in situ` >=50 & `FCS in situ` < 75 ~ "MP",
      `FCS in situ` >=25 & `FCS in situ` < 50 ~ "HP",
      `FCS in situ` < 25 ~ "UP"
    )
  )
# combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize 
DT::datatable(inTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))
}else{
    inScores <- data$fcsin |> 
  dplyr::mutate("SRS in situ" = sigfig(SRS),
                "GRS in situ" = sigfig(GRS),
                "ERS in situ" = sigfig(ERS),
                "FCS in situ" = sigfig(FCS),
                "Total occurrences in modeled area" = NA,
                "Total occurrences in protected areas" = inCounts$`Total occurrences in protected area`)|>
  dplyr::select("SRS in situ","GRS in situ","ERS in situ","FCS in situ")|>
  dplyr::mutate(
    "FCS in situ priority category" = case_when(
      `FCS in situ` >= 75 ~ "LP",
      `FCS in situ` >=50 & `FCS in situ` < 75 ~ "MP",
      `FCS in situ` >=25 & `FCS in situ` < 50 ~ "HP",
      `FCS in situ` < 25 ~ "UP"
    )
  )
# combine the data
inTables <- bind_cols(inCounts,inScores)
#visualize 
DT::datatable(inTables,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))
}



```

<br>
<br>

## Map of *in situ* conservation gap analysis measures
<br>
The SRS in situ map feature shows occurrences within the potential distribution that are outside of protected areas. Values are summarized by count per cell. The GRS in situ map feature shows the potential distribution, with distribution occurring within existing protected areas in purple and that outside of protected areas in green. The ERS in situ map features shows ecoregions within the potential distribution model that have no protected areas.There are many cases where the ecoregion area within the potential distribution model extent is very small (e.g., less than 5 cells). Utilize the “All ecoregions” layer and the ERS in situ legend to help locate these areas.


```{r echo=FALSE, message=FALSE, warning=FALSE}

if(data$NoModel == FALSE){
  uniqueSRSVals <- sort(unique(values(srs1)))

in1 <- leaflet(width = "100%")|>
  addTiles()|>
  # SRSin 
  addRasterImage(
    x = raster(srs1),
    colors = srsPal,
    group = "SRS in situ",
    project = FALSE)|>
  # protected Areas 
  addRasterImage(
    x = raster(protectedArea),
    colors = "#7570b3",
    group = "SRS in situ",
    project = FALSE )|>
  # Outline of all ecoregions
  addPolygons(
    data = allEcos,
    group = "All ecoregions",
    weight = 2,
    stroke = TRUE,
    fill = TRUE,
    fillOpacity = 0.2,
    popup = allEcos$ECO_NAME,
    color = "#60616050"
  ) |>
  leaflet::addLegend(colors = srsPal1,
            group = "SRS in situ",
            title = "SRS in situ",
            labels = c(min(uniqueSRSVals),"", "","","",max(uniqueSRSVals)))|>
  leaflet::addLegend(colors = "#7570b3",
            group = "SRS in situ",
            # title = "Protected Areas",
            labels = c("Protected Areas"))

## removing for now 2023-11-30 -- not sure what it tells that srsin does not 
# in2 <- in1 |>
#    addRasterImage(
#     x = raster(pro1),
#     colors = proColor,
#     group = "Observation within Protected Areas",
#     project = FALSE
#   )
##GRSin map 
in3 <- in1 |>
  addRasterImage(
    x = raster(grs1),
    colors = grsinColor,
    group = "GRS in situ",
    project = FALSE
  )|>
  leaflet::addLegend(colors = exPal,
            group = "GRS in situ",
            labels = c("Background area", "Potential distribution model", "Potential distribution within a protected area"))


## ERS in map 
if(class(ersinMap) != "character"){
  in3 <- in3 |>
  addRasterImage(x = raster(ersinMap),
    colors = ecoInColor,
    group = "ERS in situ",
    project = FALSE
  )|>
    leaflet::addLegend(colors = ers2$color,
              group = "ERS in situ",
              labels = ers2$ECO_NAME)
}


# add control groups 
in3 <- in3|> 
  addLayersControl(
    position = "topleft",
    options = list(collapsed = FALSE,
                   autoZIndex = FALSE),
    overlayGroups = c(
      "SRS in situ",
      "GRS in situ",
      "ERS in situ",
      "All ecoregions"))|>
  hideGroup(group = c( "SRS in situ", "ERS in situ","All ecoregions"))

in3
}else{
  "No model was generate for this species"
}

```
<br>
### *In situ* conservation ecoregions table
The table below shows all the ecoregions overlapping the potential distribution model of the taxon, indicating which do not overlap with a protected area (ERS in situ gap ecoregion = True)



The column ‘ERS in situ gap ecoregion denotes (TRUE) if an ecoregion overlaps with the potential distribution model but none of that area is within a protected area.



```{r echo=FALSE, message=FALSE, warning=FALSE}
# All ecoregions considered
# identify if they
## have modeled habitat
## are within a 50km buffer of a germplasm collection
if(data$NoModel == FALSE){
  eco3 <- data$naturalArea |>
  st_drop_geometry() |>
  dplyr::mutate(
    'ERS in situ gap ecoregion' = case_when(ECO_ID_U %in% ers2$ECO_ID_U ~ TRUE,
                                      TRUE ~ FALSE)
  ) |>
  # dplyr::left_join(misAreaIn, by = c("ECO_ID_U"= "values"))|>
  dplyr::select(
    'ERS in situ gap ecoregion',
    # 'Square KM missing' = values,
    `Unique ID` = "ECO_ID_U",
    `Ecoregion name` = "ECO_NAME",
  ) |>
  arrange(desc(`ERS in situ gap ecoregion`))
#visualize
DT::datatable(eco3,
              rownames = FALSE,
              class = "compact")
}else{
  "No model was generated for this species"
}


``` 



<br>

<br>

<br>

# Combined conservation gap analysis

This table displays all conservation gap analysis scores, providing a comparison of the ex situ and in situ scores
```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
exTable <- data$fcsex|>
  dplyr::mutate(across(c('ID'), str_replace_all, pattern = "_", replacement = " "))
names(exTable) <- c("Taxon", "Ex situ Sampling Representativeness Score", "Ex situ Geographic Representativeness Score", "Ex situ Ecological Representativeness Score", "Ex situ Final Conservation Score", "Ex situ Conservation Priority")
inTable <- data$fcsin |>
  dplyr::select(- ID)
names(inTable) <- c("In situ Sampling Representativeness Score",
                    "In situ Geographic Representativeness Score",
                    "In situ Ecological Representativeness Score",
                    "In situ Final Conservation Score",
                    "In situ Conservation Priority")
  
finalTable <- bind_cols(exTable, inTable)|>
  # round values in specific column 
  dplyr::mutate(across(c(2:5,7:10), round, digits = 2))
 


# generate the combine conservation score 
meanVal <- mean(finalTable$`Ex situ Final Conservation Score`, finalTable$`In situ Final Conservation Score`)

finalTable <- finalTable |>
  dplyr::mutate(
    `Combined Conservation Priority` = case_when(
      meanVal <= 25 ~ "UP",
      meanVal > 25 & meanVal <= 50 ~ "HP",
      meanVal > 50 & meanVal <= 75 ~ "MP",
      meanVal <= 75 ~ "UP",
    )
  )
#visualize
DT::datatable(finalTable,
              rownames = FALSE,
              class = "compact",
              options = list(dom = 't'))

```
<br>
## Conservation scores 

This figure displays all conservation gap analysis scores, providing a comparison of the ex situ and in situ scores

```{r echo=FALSE, message=FALSE, warning=FALSE,fig.align='right',out.width="100%"}
d2 <- rbind(data$fcsex, data$fcsin) %>%
  mutate(Category = c("ex situ", "in situ"))|>
  dplyr::select(
    ID,
    "SRS" = SRS,
    "GRS" = GRS,
    "ERS" = ERS,
    "FCS" = FCS,
    "Score Class" = FCS_Score,
    Category
  )

d3 <- d2 %>%
  pivot_longer(cols = c("SRS", "GRS","ERS", "FCS"), names_to = "Metric")
# define the combined score for the species 
d3[9, ] <- list(d3$ID[1], NA,"Combined conservation score","Combined Final Conservation Score",as.numeric(mean(c(d3$value[4],d3$value[8]))))

# define the combined score category 
if(d3$value[9] <= 25){
  d3$`Score Class`[9] <- "UP"
}
if(d3$value[9] > 25 & d3$value[9] <= 50){
  d3$`Score Class`[9] <- "HP"
}
if(d3$value[9] > 50 & d3$value[9] <= 75){
  d3$`Score Class`[9] <- "MP"
}
if(d3$value[9] >= 75){
  d3$`Score Class`[9] <- "LP"
}


#
d3$Metric <- factor(d3$Metric, levels = c("SRS", "GRS",
                        "ERS", "FCS", "Combined Final Conservation Score"))


plot_ly(d3,
         x = ~Metric,
         y = ~value,
         color = ~Category)%>%
  layout(xaxis = list(title = "", tickangle = 45
                      ),
         yaxis = list(title = ""),
         margin = list(b = 100),
         barmode = 'group',
         hovermode = 'compare')

```

*Note: Score values of zero will not appear on the chart.*


## Definitions of conservation gap analysis scores

### Sampling Representativeness Score (SRS)

**Ex situ**: The Sampling Representativeness Score *ex situ* (SRS ex situ) calculates the ratio of germplasm accessions (G) available in *ex situ* repositories to reference/voucher (H) records for each taxon, making use of all compiled records irrespective of whether they include coordinates.

**In situ**: The Sampling Representativeness Score *in situ* (SRS in situ) calculates the proportion of all occurrences of a taxon within its native range that fall within a protected area.

### Geographic Representativeness Score (GRS)

**Ex situ**: The Geographic Representativeness Score *ex situ* (GRS ex situ) uses 50-km-radius buffers created around each G collection coordinate point to estimate geographic areas already well collected within the potential distribution models of each taxon and then calculates the proportion of the potential distribution model covered by these buffers.

**In situ**: The Geographic Representativeness Score *in situ* (GRS in situ) compares the area (in km2) of the potential distribution model located within protected areas versus the total area of the model.

### Ecological Representativeness Score (ERS)

**Ex situ**: The Ecological Representativeness Score *ex situ* (ERS ex situ) calculates the proportion of terrestrial ecoregions represented within the G buffered areas out of the total number of ecoregions occupied by the potential distribution model. 

**In situ**: The Ecological Representativeness Score *in situ* (ERS in situ) calculates the proportion of ecoregions encompassed within the range of the taxon located inside protected areas to the ecoregions encompassed within the total area of the potential distribution model. 

### Final Conservation Score (FCS)

**Ex situ**: The Final Conservation Score *ex situ* (FCS ex situ) was derived by calculating the average of the three *ex situ* conservation metrics.

**In situ**: The Final Conservation Score *in situ* (FCS in situ) in was derived by calculating the average of the three *in situ* conservation metrics.

**FCSc-mean**: The Combined Final Conservation Score (FCSc-mean) was calculated for each taxon by averaging its final FCS ex situ and FCS in situ scores. 

Taxa were further categorized with regard to the two conservation strategies as well as in combination, with Urgent Priority (UP) for further conservation action assigned when FCS < 25, High Priority (HP) assigned when 25 ≤ FCS < 50, Medium Priority (MP) when 50 ≤ FCS < 75, and Low Priority (LP) when FCS ≥75.

## Definitions of occurrence data categories

**Occurrences**: The total number of observations and ex situ collections records for the taxon that were evaluated for use within the study.

**Occurrences with lat/long**: The total number of observations that had a valid latitude and longitude pair and were used for spatial analysis. Non valid latitude longitude pairs included but were not limited to points not on land masses, and records with only one of the two values present.

**Germplasm Records (G)**: Occurrences in which a living sample (typically a live plant or seed) is present in an *ex situ* conservation system (i.e., botanical garden, seed bank, genebank, etc.).


**Reference/voucher Records (H)**: Occurrences that have a supporting herbarium or other reference record.

**Unique Data Sources**: The number of unique database sources from which occurrences of the taxon were gathered.


